Unifying intraspecific and interspecific variation in tropical tree mortality

### 2016-06-14 ms/ms.Rnw
Added null model and made several changes to manuscript

<<preload, echo=FALSE>>=
n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
@

@@

@@

\begin{figure}
\centering
\includegraphics{figures/fig1.pdf}
\caption{Average logarithmic loss $\pm$ 95\% BCIs from 10-fold cross validation. (A) Stage 1: Comparing three functional forms of tropical tree mortality using absolute diameter growth (red) and basal area growth (blue). (B) Comparing model with and without species random effects. (C) Comparing all combinations of wood density effects.}
\label{fig1}
\end{figure}

\subsection*{Most predictive functional form}

Full model most predictive (also most biologically reasonable)
DBH DT most predictive given functional form

\subsection*{What is the most predictive combination of parameter-wood density effects?}

wood density on gamma. Technically significant on alpha too... but effect size is small (can't show this though as focus is on predictive capacity with cross val.)

\subsection*{How well does our model predict if we allow full random effects?}

Seems to account for ~20-30\% of species variability (this will be confirmed once other models are finished).

Figures:
Single figure of logloss
figure of species gamma values vs rho
response curve for low and high wood density species from dataset.

Species variability is highest on gamma (1.2) followed by alpha (0.8) and least on beta (0.3)
Mention proportion of species variability on gamma explained by wood density.

>>>>

<<preload, echo=FALSE>>=
n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))

@

@@

\begin{figure}[ht]
\centering
\includegraphics{figures/fig1.pdf}
\caption{Outline of methodology. A) Past growth used to predict future mortality. B) Three functional forms of plant mortality. C) Cross validation proceedure}
\label{fig1}
\end{figure}

@@

\subsection*{Predictive capacity of functional form and growth measure}
Using 10-fold cross validation our first aim was to determine which model form (Fig. \ref{fig1}b) and growth measure (i.e. dbh growth or basal area growth) best predicted tree mortality. We found that the model which accounted for both baseline (i.e. growth-independent) and growth-dependent hazards had a significantly lower logarithmic loss, and thus, a higher predictive capacity, relative to either the baseline only or growth-dependent only model variants (Fig. \ref{fig2}a). However, this improvement was dependent on growth measure, with predictive capacity increasing when dbh growth was used but not when basal area growth was used (Fig. \ref{fig2}a).

\begin{figure}[ht]
\includegraphics{figures/fig2.pdf}
\caption{Average logarithmic loss from 10-fold cross validation. A) Comparing various functional forms of tropical tree mortality with different growth measures different (white = no growth, grey = basal area growth, black = dbh growth) and species effects. B) Comparing model forms with various combinations of wood density effects.}
\label{fig2}
\end{figure}

\subsection*{Predictive capacity from including wood density}
Using the model with both baseline and growth-dependent hazards coupled with dbh growth we then examined: 1) whether predictive capacity increased if parameters varied as a function of a species' wood density; 2) whether the effect of wood density was predominately on growth-dependent or baseline (growth-independent) hazard parameters; and 3) what proportional gain is achieved from including wood density relative to a full species random effect model that implicitly captures all species level variation. We found that the inclusion of wood density on any combination of the three parameters significantly improved predictive performance (i.e. lowered logarithmic loss) relative to a model without wood density effects (Fig. \ref{fig2}b). Furthermore, the greatest gain in predictive accuracy occurred when we allowed the baseline hazard $\gamma$ to vary by wood density, which accounted for \Sexpr{logloss_comparisons(logloss_summaries)[['rho_v_re']]}\% of the maximum predictive gain achievable in the fall species random effect model (Fig. \ref{fig2}a).

\begin{figure}[ht]
\includegraphics{figures/fig3.pdf}
\caption{Estimated log baseline mortality against wood density. Points are individual species and solid line is the mean trendline. All errors are $\pm$ 95\% BCI.}
\label{fig3}
\end{figure}

\subsection*{Effects of wood density and time}
Once determining that the most predictive model was one that accounted for both growth-dependent and baseline hazards, used dbh growth, and allowed $\gamma$ to vary by wood density, we then fitted two final models to the entire dataset. The first contained both species random effects on all parameters and wood density on the baseline hazard, $\gamma$. This allowed us to examine the range of species growth-mortality curves, the effect of wood density and how mortality varied over time. We also fitted a second model that only contained species random effects. This allowed us to determine which parameter varied most by species and what proportion of the baseline hazard species variability was explained by wood density. We found that baseline hazard rates were significantly lower for species with high wood density relative to those with low wood density (Fig. \ref{fig3}). That is, high wood density species had a lower probability of dying irrespective of growth rate. For example, our model predicts that on average a species with a wood density of 0.3 has a instantaneous mortality rate of XX, whereas a species with a wood density of 0.8 has a instantenaous mortality rate of XX, translating to an annual mortality rate of XX\% and XX\%, respectively. Furthermore, our model also revealed that instantaneous mortality rates varied over time. Specifically, mortality rates at BCI decreased over time with hazard rates being scaled by \Sexpr{census_effects(final_model,1)},\Sexpr{census_effects(final_model,2)} and \Sexpr{census_effects(final_model,3)} for census periods, 1995\textendash 2000, 2000\textendash 2005 and 2005\textendash 2010, respectively.

\begin{figure}[ht]
\includegraphics{figures/fig4.pdf}[ht]
\caption{Estimated curves of instantaneous mortality rate and annual mortality probability. A,C) Highlight expected response curves for a hypothetical median species at low (0.3) and high (0.8) wood density. B,D) Predicted response curves for all species. Errors are $\pm$ 95\% BCI.}
\label{fig4}
\end{figure}

\subsection*{Sources of variation}
We found that the baseline hazard, $\gamma$, varied the most between species followed by $\alpha$ and least on $\beta$. We also found that the wood density effect on $\gamma$ explained approximately XX\% of the species variability on the baseline hazard.
